---
title: "rgbif"
author: "LH"
date: "2023-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Packages
```{r}
library(here)
library(rgbif)
library(sf)
library(maps)
library(measurements)
library(tidyverse)
library(GGally)
library(ggridges)
library(lubridate)
library(kableExtra)
library(RColorBrewer)
```

# Import bird data for SMF area

Process the bounding box for SMF

top left:
 lat: 34.465504째
 long: -119.771787째
 
bottom right: 
  lat:  34.454424째
  long: -119.750857째
  
c(lng_min, lat_min, lng_max, lat_max)
st_bbox(xmin, ymin, xmax and ymax)
c(-119.750857, 34.454424, 119.771787, 34.465504)

For converting shp to wkt:
https://www.erikkusch.com/courses/gbif/datadiscovery/

Test your WKT object here: http://arthur-e.github.io/Wicket/sandbox-gmaps3.html
(Print your wkt object n)

```{r}
SMF_wkt <- gbif_bbox2wkt(maxx = -119.750857, miny = 34.454424, minx = -119.771787, maxy = 34.465504)
# SMF_wkt # print this, open the sand box link, copy the results into the box, then press Map It!
```

Import GBIF data for R
Class aves (birds) is taxon ID 212: https://www.gbif.org/species/212
```{r}
# download GBIF occurrence data for this species; this takes time if there are many data points!
gbif_data <- occ_data(hasCoordinate = TRUE, geometry = SMF_wkt, classKey = 212, year = "1990, 2023", limit = 30000)

# take a look at the downloaded data:
gbif_data

# if "Records found" is larger than "Records returned", you need to increase the 'limit' argument above -- see help(occ_data) for options and limitations
# as of 2023-11, there are ~14k records for birds in our region

# get the DOIs for citing these data properly
citation <- gbif_citation(gbif_data)

# get the columns that matter for mapping and cleaning the occurrence data
myspecies_coords <- gbif_data$data[ , c("decimalLongitude", "decimalLatitude", "order", "family", "genus", "species", "taxonomicStatus", "iucnRedListCategory", "individualCount", "eventDate", "year", "month", "day", "eventTime", "type", "recordedBy", "coordinateUncertaintyInMeters", "collectionCode", "references")]
#head(myspecies_coords)

# remove NA
myspecies_coords <- myspecies_coords %>% 
  filter(is.na(species) ==FALSE)

```

Test data and display

```{r, eval = FALSE}
# convert frame points to decimal degrees
lat_minx <- -119.771787
long_maxy <- 34.465504
lat_maxx <- -119.750857
long_miny <- 34.454424

# map the occurrence data to test it
# if the map doesn't appear right at first, run this command again
map("world", xlim = c(lat_minx, lat_maxx), ylim = c(long_miny, long_maxy)) # set boundaries ("frame")
points(myspecies_coords[ , c("decimalLongitude", "decimalLatitude")], pch = ".") # add points
```

Test synonyms
```{r}
synonym_data <- myspecies_coords[myspecies_coords$taxonomicStatus == "SYNONYM",]

# all species marked as synonyms
unique(synonym_data$species)

# the synonyms here don't seem to be species with two scientific names
# we will leave them in, and not filter them out for analyses
# Haemorhous mexicanus is sometimes called synonym and sometimes not, so it was only partially removed while filtering in previous versions

```

# Summary statistics and visualization

Summary statistics
```{r}
# individualCount column is the # of individuals per observation

# some observations say NA, and we will assume NA = 1
# head(myspecies_coords[is.na(myspecies_coords$individualCount) == TRUE,]) <- 1 # test to see how many are NA (returned 675)
myspecies_coords[is.na(myspecies_coords$individualCount) == TRUE,]$individualCount <- 1
# test that it worked (should return 1, not NA)
# min(myspecies_coords$individualCount, na.rm = FALSE)

myspecies_coords$month_abb <- factor(month.abb[myspecies_coords$month],levels=month.abb) # add month abbreviation
```

Create list of unique species and counts
```{r}
# list of unique species and their classification
alltaxa <- myspecies_coords %>% 
  distinct(order, family, genus, species) %>%  # keep only this data
  filter(is.na(species) == FALSE) # remove NA

# counts for each species
sum_spp <- myspecies_coords %>% 
  group_by(species) %>% 
  summarise(total = length(species))

```

## By year and order

```{r}
sum_order_year <- myspecies_coords %>% 
  group_by(order, year) %>% 
  summarise(total = sum(individualCount))
```

Visualize
```{r}
# bird order by year
ggplot(sum_order_year, aes(fill=order, y=total, x=year)) + 
    geom_bar(stat="identity")

# ridgeline plot
ggplot(myspecies_coords, aes(x = year, y = order, fill = order)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none") +
  ylab("Order") +
  xlab("Year")

```

## By month and order

```{r}
sum_order_month <- myspecies_coords %>% 
  filter(year > 2009 & year < 2023) %>% # keep only records from 2010 and on and before 2023 because it gets cut off
  group_by(order, month) %>% 
  summarise(total = sum(individualCount))

sum_order_month$month_abb <- factor(month.abb[sum_order_month$month],levels=month.abb) # add month abbreviation

```

Visualize 
```{r}
# bird order by month
ggplot(sum_order_month, aes(fill=order, y=total, x=month_abb)) + 
  geom_bar(stat="identity", position="fill") 

# bird order by month
ggplot(sum_order_month, aes(fill=order, y=total, x=month_abb)) + 
  geom_bar(stat="identity") 

# bird order by month - all bars same height (percent stacked bar chart), but faceted
ggplot(sum_order_month, aes(fill=order, y=total, x=month_abb)) + 
  geom_bar(stat="identity") +
  facet_wrap(~order) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

# ridgeline plot
ggplot(myspecies_coords, aes(x = month, y = order, fill = order)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none") +
  ylab("Order") +
  xlab("Month")

```

## Add in data about diet

Export csv to fill in guilds/ecology
```{r, eval = FALSE}
write.csv(alltaxa, file = here("data", "output", "alltaxa.csv"))
```

Import new data
```{r}
alltaxa_diet <- read.csv(here("data", "input", "alltaxa_diet.csv"))
```

bind new categories to bird species, then to all observations
```{r}
dietkey <- alltaxa_diet %>% 
  select(species, diet) 

myspecies_coords <- left_join(myspecies_coords, dietkey, by = "species")
```

group_by and summarize
```{r}
sum_diet_month <- myspecies_coords %>% 
  filter(year > 2009 & year < 2023) %>% # keep only records from 2010 and on and before 2023 (because cuts off partway through 2023)
  filter(is.na(species) == FALSE) %>% 
  group_by(diet, month) %>% 
  summarise(total = sum(individualCount))

sum_diet_month$month_abb <- factor(month.abb[sum_diet_month$month],levels=month.abb) # add month abbreviation
```

visualize
```{r}
# bird diet by month
ggplot(sum_diet_month, aes(fill=diet, y=total, x=month_abb)) + 
  geom_bar(stat="identity") 

# bird diet by month - all bars same height (percent stacked bar chart)
gg_diet_m_stacked <- ggplot(sum_diet_month, aes(fill=diet, y=total, x=month_abb)) + 
  geom_bar(stat="identity", position="fill") +
  ylab("Diet (proportion)") +
  xlab("Month") +
  labs(fill = "Diet")

gg_diet_m_stacked

# bird diet by month - all bars same height (percent stacked bar chart), but faceted
ggplot(sum_diet_month, aes(fill=diet, y=total, x=month_abb)) + 
  geom_bar(stat="identity") +
  facet_wrap(~diet) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

# ridgeline plot
ggplot(myspecies_coords, aes(x = month, y = diet, fill = diet)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none") +
  ylab("Diet") +
  xlab("Month")

```

## Sampling Effort

Question: How does sampling effort effect these results?

Count up unique name-date combinations. For every person that made observations on a day, that is 1 sampling effort for that day. 
recordedBy = observer
eventDate (truncated to year-month-day) = day
type = PhysicalObject means museum specimen

```{r}
# count NAs in recordedBy (we want very few)
sum(is.na(myspecies_coords$recordedBy))

# remove all observations except for ebird (sampling effort varies between platforms/methods)
myspecies_ebird <- myspecies_coords[myspecies_coords$collectionCode == "EBIRD",]

# find observations per person to see if there are anomalies (note ebird observations will be high)
ebird_sumstats <- myspecies_ebird %>% 
  group_by(recordedBy, month) %>% 
  summarize(n()) 

ebird_sumstats_m <- ebird_sumstats %>% 
  group_by(month) %>% 
  summarize(count = n())

# graph
gg_effort <- ggplot(ebird_sumstats_m, aes(x = month, y = count)) +
  geom_col() +
  ylab("Sampling Events") +
  xlab("Month")

gg_effort
```

Now, let's compare the number of observations to the level of effort to see how correlated they are

All combined (no details)
```{r}
# rename datasets and columns
ebird_effort <- ebird_sumstats_m 
ebird_effort$effort <- ebird_effort$count

# keep month and effort columns
effort_combine <- ebird_effort %>% 
  select(month, effort) 
# remove some columns
effort_combine$order <- NULL
sum_order_month$month_abb <- NULL

effort_combine$month_abb <- factor(month.abb[effort_combine$month],levels=month.abb) # add month abbreviation

# summarize ebird data (by number of observations, not effort)
sum_month <- myspecies_ebird %>% 
  group_by(month_abb) %>% 
  summarise(observations = sum(individualCount))

# combine both
month_effort <- full_join(sum_month, effort_combine, by = c("month_abb"))

# interpolate (to make next plot nicer, or don't bother)
# use this: https://stackoverflow.com/questions/45227527/ggplot2-cannot-color-area-between-intersecting-lines-using-geom-ribbon

```

Plot and highlight diversions between effort and observations
```{r}
ggplot(data = month_effort) +
  geom_point(aes(x = month_abb, y = effort, color = "Sampling Effort")) +
  geom_line(aes(x = month_abb, y = effort, group = 1)) +
  geom_point(aes(x = month_abb, y = observations, color = "Observed")) +
  geom_line(aes(x = month_abb, y = observations, group = 1)) +
  geom_ribbon(aes(x = month_abb, ymin = effort, ymax = pmin(observations, effort)), fill = "red", group=1, alpha=0.3) +
  geom_ribbon(aes(x = month_abb, ymin = observations, ymax = pmin(effort, observations)), fill = "green", group=1, alpha=0.3) +
  xlab("Month") +
  ylab("Sampling Effort")

```

For Order and sampling effort
```{r, eval = FALSE}
ebird_effort <- ebird_sumstats_m 
ebird_effort$total <- ebird_effort$count*100

effort_combine <- ebird_effort %>% 
  select(month, total) 

effort_combine$order <- NA
sum_order_month$month_abb <- NULL
effort_combine$tag <- "Effort"
sum_order_month$tag <- "a"

combine <- full_join(sum_order_month, effort_combine, by = c("month", "total", "order", "tag"))

combine$month_abb <- factor(month.abb[combine$month],levels=month.abb) # add month abbreviation

```

Bar plot of observations (by order) next to to effort
```{r, eval = FALSE}
ggplot() +
  geom_col(data = filter(combine, tag == "a"), aes(y = total, x = month_abb, fill = order), width = .3, just = 1) +
  geom_bar(data = filter(combine, tag == "Effort"), aes(x = month_abb, y = total, color = tag), stat='identity', width = .3, just = 0) +
  theme_bw() +
  xlab("Month") +
  ylab("Total Observations") +
  labs(fill = "Order", color = "Relative Sampling Effort") 

```

Sampling effort ratio and plots
```{r}
month_effort$ratio <- month_effort$observations/month_effort$effort # calculate ratio
effort_mean <- mean(month_effort$ratio) # ratio = observations per sampling event

# plot

# ratio
ggplot(month_effort) +
  geom_point(aes(x = month_abb, y = ratio)) +
  geom_line(aes(x = month_abb, y = ratio), group = 1) +
  theme_bw() +
  xlab("Month") +
  ylab("Observations per Sampling Event (ratio)") +
  ylim(c(0, NA)) +
  geom_hline(yintercept = effort_mean, color = "blue")

# effort
ggplot(month_effort) +
  geom_point(aes(x = month_abb, y = effort)) +
  geom_line(aes(x = month_abb, y = effort), group = 1) +
  theme_bw() +
  xlab("Month") +
  ylab("Number of Sampling Events (effort)") 

# observations
ggplot(month_effort) +
  geom_point(aes(x = month_abb, y = observations)) +
  geom_line(aes(x = month_abb, y = observations), group = 1) +
  theme_bw() +
  xlab("Month") +
  ylab("Number of Observations") 

```

Now, try looking at month-years, instead of months aggregated of years, and then get the mean ratio of those (to account for differences between years)
```{r}
# find observations per person
ebird_sumstats_m_y <- myspecies_ebird %>% 
  group_by(recordedBy, month, year) %>% 
  summarize(n()) 

# count up effort
ebird_sumstats_m_y <- ebird_sumstats_m_y %>% 
  group_by(month, year) %>% 
  summarize(effort = n())

# add month abbreviation
ebird_sumstats_m_y$month_abb <- factor(month.abb[ebird_sumstats_m_y$month],levels=month.abb) 

# summarize ebird data (by number of observations, not effort)
sum_m_y <- myspecies_ebird %>% 
  group_by(month_abb, year) %>% 
  summarise(observations = sum(individualCount))

# combine both
ratio_m_y <- full_join(sum_m_y, ebird_sumstats_m_y, by = c("month_abb", "year"))

# take ratio (observations/effort)
ratio_m_y$ratio <- ratio_m_y$observations/ratio_m_y$effort 

ratio_m_y <- ratio_m_y %>% 
  filter(year > 2010) 
```

```{r}
# plot with all lines together
ggplot(ratio_m_y) +
  geom_point(aes(x = month_abb, y = ratio, color = year, group = year)) +
  geom_line(aes(x = month_abb, y = ratio, color = year, group = year)) +
  theme_bw() +
  xlab("Month") +
  ylab("Number of Observations") +
  scale_color_distiller(palette = "Spectral")

# plot with one year per panel
ggplot(ratio_m_y) +
  geom_point(aes(x = month, y = ratio)) +
  geom_line(aes(x = month, y = ratio)) +
  theme_bw() +
  xlab("Month") +
  ylab("Number of Observations") +
  scale_color_distiller(palette = "Spectral") +
  facet_wrap(~year)

# plot with one month per panel
ggplot(ratio_m_y) +
  geom_point(aes(x = year, y = ratio)) +
  geom_line(aes(x = year, y = ratio)) +
  theme_bw() +
  xlab("Year") +
  ylab("Number of Observations") +
  scale_color_distiller(palette = "Spectral") +
  facet_wrap(~month_abb) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

Take the mean ratio of those month-years
```{r}
ratio_m_avg <- ratio_m_y %>% 
  group_by(month) %>% 
  summarize(avg_ratio = mean(ratio))

# add month abbreviation
ratio_m_avg$month_abb <- factor(month.abb[ratio_m_avg$month],levels=month.abb) 

ratio_mean_grain <- mean(ratio_m_avg$avg_ratio)

# ratio
ggplot(ratio_m_avg) +
  geom_point(aes(x = month_abb, y = avg_ratio)) +
  geom_line(aes(x = month_abb, y = avg_ratio), group = 1) +
  theme_bw() +
  xlab("Month") +
  ylab("Observations per Sampling Event (high-grain ratio)") +
  ylim(c(0, NA)) +
  geom_hline(yintercept = ratio_mean_grain, color = "blue")

```

## Rare species present at SMF

```{r}
rarespp <- myspecies_coords %>% 
  filter(iucnRedListCategory != "LC" & iucnRedListCategory != "NE")

unique(rarespp$species)
```

## Lanius ludovicianus - Loggerhead Shrike

```{r}
# filter observations for only this sp.
losh <- myspecies_coords %>% 
  filter(species == "Lanius ludovicianus")

# seasonality of loggerhead shrikes (shrikes per month)
losh_m <- losh %>% 
  group_by(month) %>% 
  summarize(total = sum(individualCount)) 

losh_m$month_abb <- factor(month.abb[losh_m$month],levels=month.abb) # add month abbreviation

# plot
ggplot() +
  geom_col(data = losh_m, aes(y = total, x = month_abb)) +
  scale_x_discrete(limits = month.abb) + # add in empty months
  xlab("Month") +
  ylab("Loggerhead shrike observations")

```

# Species richness 

Species richness is a component of biodiversity. It means the total number of species, with no regard for the proportionality of species. 

```{r}
# total richness - all species, all years (no NAs)
length(unique(myspecies_coords$species[is.na(myspecies_coords$species) == FALSE]))

# richness per year
richness_y <- myspecies_coords %>% 
  filter(is.na(species) == FALSE) %>% # remove NAs
  select(species, year) %>% 
  group_by(year, species) %>% 
  filter(!duplicated(species, year)) %>% # remove duplicate species per year
  group_by(year) %>% 
  summarize(n = length(year))

# richness per month
richness_m <- myspecies_coords %>% 
  filter(is.na(species) == FALSE) %>% # remove NAs
  select(species, month) %>% 
  group_by(month, species) %>% 
  filter(!duplicated(species, month)) %>% # remove duplicate species per year
  group_by(month) %>% 
  summarize(n = length(month))

richness_m$month_abb <- factor(month.abb[richness_m$month],levels=month.abb) # add month abbreviation

```

Plots for richness
```{r}
# richness by month
ggplot() +
  geom_col(data = richness_m, aes(y = n, x = month_abb)) +
  scale_x_discrete(limits = month.abb) + # add in empty months
  xlab("Month") +
  ylab("Richness")

# richness by year
ggplot() +
  geom_col(data = richness_y, aes(y = n, x = year)) +
  xlab("Year") +
  ylab("Richness")
```

Richness by order
```{r}
# richness per order
richness_order <- myspecies_coords %>% 
  filter(is.na(species) == FALSE) %>% # remove NAs 
  select(species, order) %>% 
  group_by(order, species) %>% 
  filter(!duplicated(species, order)) %>% # remove duplicate species per year
  group_by(order) %>% 
  summarize(n = length(order)) %>% 
  mutate(n_perc = 100*n/sum(n))

# we can also add in the proportion of individuals which are in those orders (2010-2022 data): columns are order and proportion
sum_order <- sum_order_month %>% 
  group_by(order) %>% 
  summarize(proportion_m_avg = mean(total), proportion_y_sum = sum(total)) %>% 
  mutate(proportion_m_avg_perc = 100*proportion_m_avg/sum(proportion_m_avg),
         proportion_y_sum_perc = 100*proportion_y_sum/sum(proportion_y_sum))

# join
sum_order_perc <- full_join(richness_order, sum_order, by = "order") %>% 
  select(order, n, n_perc, proportion_y_sum_perc) 

# rename for formatting
names(sum_order_perc) <- c("Order", "Total Species Richness", "Proportion of Richness", "Proportion of Observations")

# print table with info on proportion of individuals in order
sum_order_perc %>%
  kbl(digits = 1) %>%
  kable_minimal()

```

Richness by diet
```{r}
# richness per diet
richness_diet <- myspecies_coords %>% 
  filter(is.na(species) == FALSE) %>% # remove NAs
  select(species, diet) %>% 
  group_by(diet, species) %>% 
  filter(!duplicated(species, diet)) %>% # remove duplicate species per year
  group_by(diet) %>% 
  summarize(n = length(diet)) %>% 
  mutate(n_perc = 100*n/sum(n))

# we can also add in the proportion of individuals which are in those diets (2010-2022 data): columns are diet and proportion
sum_diet <- sum_diet_month %>% 
  group_by(diet) %>% 
  summarize(proportion_m_avg = mean(total), proportion_y_sum = sum(total)) %>% 
  mutate(proportion_m_avg_perc = 100*proportion_m_avg/sum(proportion_m_avg),
         proportion_y_sum_perc = 100*proportion_y_sum/sum(proportion_y_sum))

# join
sum_diet_perc <- full_join(richness_diet, sum_diet, by = "diet") %>% 
  select(diet, n, n_perc, proportion_y_sum_perc) 

# rename for formatting
names(sum_diet_perc) <- c("Diet", "Total Species Richness", "Proportion of Richness", "Proportion of Observations")

# print table with info on proportion of individuals in diet
sum_diet_perc %>%
  kbl(digits = 1) %>%
  kable_minimal()

```

# Santa Barbara Breeding Bird Study

Read in data (https://ucsb.maps.arcgis.com/apps/webappviewer/index.html?id=d5479baf24dc498a8e8f3894f298ed2d)
Main site: https://santabarbaraaudubon.org/santa-barbara-county-breeding-bird-study/ 
Shout out to Mark Holmgren for this amazing dataset!
```{r}
sbbbs <- read.csv(here("data", "input", "SBBBS_SMF_2023-12.csv"))

# common names/sci names from earlier
all_diet_common <- read.csv(here("data", "output", "alltaxa_diet_commonname.csv"))

```

Summary statistics
```{r}
# colnames(sbbbs) # so we know what column names we are working with
# rename some that look funny
colnames(sbbbs)[4] <- "Species"
sbbbs$Common.Name <- str_to_sentence(sbbbs$Common.Name) # first word uppercase, second lowercase
# the only column in common with gbif data is Species

# stats by species
sbbbs_stats <- sbbbs %>% 
  group_by(Species, Common.Name, Code) %>% 
  summarize(total = length(Species))

# stats by date
sbbbs_time <- sbbbs
# convert column to date class
sbbbs_time$Observation.Date <- as.Date(sbbbs$Observation.Date, format = "%d-%b-%y") 

# by year
# create year column
sbbbs_time$year <- format(as.Date(sbbbs_time$Observation.Date),"%Y") 
# summary by year
sbbbs_stats_y <- sbbbs_time %>% 
  group_by(year) %>% 
  summarize(Total = length(year))
# this looks super variable
# we can only conclude that sampling effort varies wildly

# by month
# create month column
sbbbs_time$month <- as.numeric(format(as.Date(sbbbs_time$Observation.Date),"%m")) 
# summary by month
sbbbs_stats_m <- sbbbs_time %>% 
  group_by(month) %>% 
  summarize(Total = length(month))
# april, may, and june are most popular months for breeding and nesting birds

```

# Edit existing tables 

Combine spp list with species from SMF bird list (that were not already accounted for)
```{r}
smfwebsite_spp <- read.csv(here("data", "input", "species_diff_from_smf_website.csv"))
```

Add species counts to spp list (run as needed)
```{r, eval = FALSE}
# rename columns for join
colnames(sum_spp)[1] <- "Species"
colnames(sum_spp)[2] <- "Total.Observed"
# do join of spp and spp counts
all_diet_common_sum <- full_join(all_diet_common, sum_spp, by = "Species")

```

# Export tables/graphs/data

Tables
```{r, eval = FALSE}

# diet & number of species
write.csv(sum_diet_perc, file = here("data", "output", "sum_diet_perc.csv"))

# order & number of species
write.csv(sum_order_perc, file = here("data", "output", "sum_order_perc.csv"))

# export all species & diets (to later add in common names)
write.csv(alltaxa_diet, file = here("data", "output", "alltaxa_diet.csv"))

# breeding birds per month
write.csv(sbbbs_stats_m, file = here("data", "output", "sbbbs_stats_m.csv"))

# breeding birds by species
write.csv(sbbbs_stats, file = here("data", "output", "sbbbs_stats_spp.csv"))

# spp and spp counts
write.csv(all_diet_common_sum, file = here("data", "output", "all_diet_common_sum.csv"))
```

Graphs
```{r}
ggsave(path = here("data", "output"), plot = gg_diet_m_stacked, filename = "gg_diet_m_stacked.jpg")

gg_diet_m_stacked
```

